
```{r global}

library(pacman)
p_load(papaja, beeswarm, ggplot2, dplyr, readxl, tinytex, emmeans, tidyr, afex, wesanderson, plotly, htmlwidgets, magick, ggridges, purrr, forcats, viridis, pals, paletteer)

set.seed(123)

#### data wrangling ####

speciesList <- c('bre', 'mil', 'mul', 'oce', 'orn', 'pul')

# read data
dat <- read.csv("PrettyDat.csv")


#expand dat into set of discrete time series by trial_id, sequence_id, and fish_id
expanded <- dat %>%
  arrange(trial_id, sequence_id, order_id, fish_id, species) %>%
  group_by(trial_id, sequence_id, fish_id, species) %>%
  summarise(
    series = list(rep(state, length)),
    .groups = "drop"
  )

### conditional length calculations

# add next state and length column for easier indexing
dat2 <- dat %>%
  arrange(trial_id, sequence_id, order_id, fish_id, species) %>%
  group_by(trial_id, sequence_id, fish_id, species) %>%
  mutate(
    next_state  = lead(state),
    next_length = lead(length)
  ) %>%
  ungroup() %>%
  filter(!is.na(next_state))

#calculate average conditional length of state
lengthConds <- dat2 %>%
  group_by(state, next_state, species) %>%
  summarise(mean_next_length = mean(next_length), .groups = "drop")

# format for plotting
lengthConds <- lengthConds %>%
  mutate(
    state = factor(state),
    next_state = factor(next_state),
    #x = as.numeric(state),
    #y = as.numeric(next_state),
    #x_offset = ifelse(fish_id == "dom", -0.25, 0.25)
  )

### conditional state distributions ###

# duplicate stagger sequence by 1 to get prev/next state pairings at every timepoint t
get.transitions <- function(seq) {
  if (length(seq) < 2) {
    return(NULL)
  }
  data.frame(
    prev = head(seq, -1),
    nex = tail(seq, -1),
    stringsAsFactors = FALSE
  )
}

fullProbs <- NULL

# filtered version without same-state transitions
fullProbsF <- NULL

for (specie in speciesList) { #for each species:
  
  #subset
  sExpanded <- expanded[expanded$species==specie,]
  
  #get transitions
  sTransitions <- lapply(sExpanded$series, get.transitions)
  sTransitions <- bind_rows(sTransitions, .id = "id")
  
  #for filtered version, remove same-state transitions
  sTransitionsF <- sTransitions %>%
  filter(prev != nex)
  
  #proportion of transitions
  sProbs <- sTransitions %>%
    group_by(id, prev, nex) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(id, prev) %>%
    mutate(prob = n / sum(n)) %>%
    ungroup()
  
  #filtered
  sProbsF <- sTransitionsF %>%
    group_by(id, prev, nex) %>%
    summarise(n = n(), .groups = "drop") %>%
    group_by(id, prev) %>%
    mutate(prob = n / sum(n)) %>%
    ungroup()
  
  
  sProbs$species = specie
  sProbsF$species = specie
  
  fullProbs <- rbind(fullProbs, sProbs)
  fullProbsF <- rbind(fullProbsF, sProbsF)
  
}

```

# Useful Numbers

```{r nums}

speciesList <- c('bre', 'mil', 'mul', 'oce', 'orn', 'pul')

for (s in speciesList){
  
  temp <- expanded[expanded$species==s,]
  
  # calcs
  nSeq <- nrow(temp)
  nInts <- nSeq/2
  seqQuant75 = quantile(nchar(temp$series), probs = .75)
  seqQuant50 = quantile(nchar(temp$series), probs = .50)
  meanL <- mean(nchar(temp$series))
  minL <- min(nchar(temp$series))
  maxL <- max(nchar(temp$series))
  
  cat(
  s, ":\n",
  "Number of Sequences:", nSeq, "\n",
  "Number of Interactions:", nInts, "\n",
  "Min Sequence Length:", minL, "\n",
  "Max Sequence Length:", maxL, "\n",
  "Sequence Length (mode):", seqQuant50, "\n", 
  "Sequence Length (upper quantile):", seqQuant75, "\n\n"
)

}

# sequence length histogram (also shows count)
ggplot(expanded, aes(x = nchar(series), fill = species, group = species)) +
  geom_histogram(bins = 20)+
  facet_wrap(~species, scales = "free_y")+
  scale_fill_paletteer_d("rcartocolor::Prism", guide = "none")+
  scale_x_log10(breaks = c(100, 1000, 10000, 100000), labels = c('1e2', '1e3', '1e4', '1e5'))+
  labs(x = "Length of Sequence", y = "Count")+
  theme_apa()

```

# Initial Analyses

Understanding the structure of fish posture first requires knowledge of:

1. how long each state lasts: average length
2. how long one state lasts, conditional on previous state(s): conditional average length
3. how often each state occurs: state distributions
4. what state comes next, conditional on previous state: conditional state distributions
5. the extent of memory within the system: tau-lagged eigenvector decay
6. whether state probabilities shift dynamically: transition probabilities as a function of lag

## 1. Average State Length

```{r 1}

ggplot(dat, aes(x = state, y = length, fill = species, group = species)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge")+
  facet_wrap(~species)+
  scale_fill_paletteer_d("rcartocolor::Prism", guide = "none")+
  labs(y = "Length of State", x = "State")+
  theme_apa()

```

## 2. Conditional Average Next State Length

```{r 2}

ggplot(lengthConds,
       aes(state, next_state, fill = log10(mean_next_length))) +
  geom_tile() +
  facet_wrap(~ species) +
  scale_fill_paletteer_c("ggthemes::Red-Gold")+
  labs(fill = "Mean Length (Log10)", y = "Next State", x = "Previous State")+
  coord_equal() +
  theme_minimal()+
  theme(panel.grid = element_line(color = "gray85", size = 0.25,linetype = 2), 
        panel.spacing = unit(3, "lines"))

```

## 3. State Distributions

```{r 3}

# ggplot(dat, aes(x = state, y = after_stat(prop), fill = species, group = species)) +
#   geom_bar(stat = "count", position = "dodge", color = NA) +
#   scale_fill_paletteer_d("rcartocolor::Prism", name = "Species")+
#   labs(y = "Proportion Time Spent in State", x = "State")+
#   theme_apa()

state_time <- dat %>%
  group_by(species, state) %>%
  summarise(total_length = sum(length), .groups = "drop") %>%
  group_by(species) %>%
  mutate(prop_time = total_length / sum(total_length)) %>%
  ungroup()

ggplot(state_time, aes(x = state, y = prop_time, fill = species, group = species)) +
  facet_wrap(~species)+
  geom_bar(stat = "identity", position = "dodge", color = NA) +
  scale_fill_paletteer_d("rcartocolor::Prism", guide = "none") +
  labs(y = "Proportion Time Spent in State", x = "State") +
  theme_apa()

```

## 4. Conditional State Distributions

```{r 4}

# unfiltered

# ggplot(fullProbs, aes(x = prev, y = n, fill = fct_rev(factor(nex, levels = sort(unique(nex)))))) +
#   geom_bar(stat = "identity", position = "fill") +
#   facet_wrap(~ fish, labeller = as_labeller(c(dom = "Dom", sub = "Sub")))+
#   labs(x = "Previous State", y = "Proportion of Transitions", fill = "Next State") +
#   scale_fill_paletteer_d("rcartocolor::Prism")+
#   theme_minimal()



# filtered
ggplot(fullProbsF, aes(x = prev, y = n, fill = fct_rev(factor(nex, levels = sort(unique(nex)))))) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~ species)+
  labs(x = "Previous State", y = "Proportion of Transitions", fill = "Next State") +
  scale_fill_paletteer_d("rcartocolor::Prism")+
  theme_minimal()


# list of unique combos
fullProbsF %>%
  group_by(species, prev) %>%
  summarise(nex_set = paste(sort(unique(nex)), collapse = ", "),
            .groups = "drop") %>%
  group_split(species) %>%
  walk(function(x) {
    cat("Species:", unique(x$species), "\n")
    
    apply(x, 1, function(row) {
      cat("Prev =", row["prev"], ": Nexts =", row["nex_set"], "\n")
    })
    
    cat("\n")
  })

```

## 5. Memory within the System (Eigenvalue Decay Rate)

```{r}
lagT <- function(seqs, tau) {
  
  states <- sort(unique(unlist(seqs)))
  n <- length(states)
  state_idx <- setNames(seq_along(states), states)
  
  mat <- matrix(0, nrow=n, ncol=n, dimnames=list(states, states))
  
  for (seq in seqs) {
    if (length(seq) <= tau) next
    
    #create transition matrix at lag tau
    for (i in 1:(length(seq) - tau)) {
      from <- as.character(seq[i])
      to   <- as.character(seq[i + tau])
      mat[from, to] <- mat[from, to] + 1
    }
  }
  
  #row-normalize (rows = "from" states)
  row_sums <- rowSums(mat)
  mat <- mat / ifelse(row_sums > 0, row_sums, 1)
  return(mat)
}

#first order markov tau is just T(1)^tau
matrixPower <- function(M, tau) {
  result <- diag(nrow(M))
  base <- M
  while (tau > 0) {
    if (tau %% 2 == 1) result <- result %*% base
    base <- base %*% base
    tau <- tau %/% 2
  }
  return(result)
}

#### visualizations ####

tau_values <- c(1, 50, 100, 200, 500, 1000, 2500, 5000, 7500, 10000, 15000) #upper quantile of sequence length is ~10,000 for each species
n_modes <- 5  # plot top 5 non-trivial eigenvalues (mu = 2->6 in paper)

speciesList <- c('bre', 'mil', 'mul', 'oce', 'orn', 'pul')

colors <- paletteer_d("rcartocolor::Prism")
mycols <- colors[4:9]


for (specie in speciesList) {
  dat3 <- dat %>%
    filter(species == specie) %>%
    arrange(trial_id, sequence_id, fish_id, order_id)
  
  trueSeqs <- dat3 %>%
    group_by(trial_id, sequence_id, fish_id) %>%
    summarise(series = list(as.character(rep(state, length))), .groups = "drop") %>%
    pull(series)
  
  # Q3 of sequence lengths
  seq_lengths <- sapply(trueSeqs, length)
  q3_len      <- quantile(seq_lengths, 0.75)
  
  #T(1) empirical
  P1 <- lagT(trueSeqs, tau = 1)
  
  #init
  eig_empirical <- matrix(NA, nrow=length(tau_values), ncol=n_modes)
  eig_markov    <- matrix(NA, nrow=length(tau_values), ncol=n_modes)
  
  for (k in seq_along(tau_values)) {
    tau <- tau_values[k]
    
    Ptau <- lagT(trueSeqs, tau)
    
    ev_emp <- sort(Mod(eigen(Ptau, only.values=TRUE)$values), decreasing=TRUE)
    eig_empirical[k, ] <- ev_emp[2:(n_modes+1)]
    
    ev_1 <- sort(Mod(eigen(P1, only.values=TRUE)$values), decreasing=TRUE)
    eig_markov[k, ] <- ev_1[2:(n_modes+1)]^tau
  }
  
  # Build tidy long-format data frames
  mode_labels <- paste0("i==", 2:(n_modes+1))
  
  df_emp <- as.data.frame(eig_empirical) %>%
    setNames(mode_labels) %>%
    mutate(tau = tau_values) %>%
    pivot_longer(-tau, names_to="mode", values_to="lambda") %>%
    mutate(type = "Empirical")
  
  df_mkv <- as.data.frame(eig_markov) %>%
    setNames(mode_labels) %>%
    mutate(tau = tau_values) %>%
    pivot_longer(-tau, names_to="mode", values_to="lambda") %>%
    mutate(type = "Markov prediction")
  
  df_plot <- bind_rows(df_emp, df_mkv)
  
  max_tau <- max(tau_values)
  
  p <- ggplot(df_plot, aes(x=tau, y=lambda, color=mode, linetype=type, group=interaction(mode, type))) +
    geom_line(linewidth=0.8) +
    geom_point(data = filter(df_plot, type == "Empirical"), size=2) +
    # Q3 shaded overlay
    annotate("rect",
             xmin = q3_len, xmax = max_tau,
             ymin = 0,      ymax = 1,
             fill = "grey", alpha = 0.45) +
    annotate("segment",
             x = q3_len, xend = q3_len, y = 0, yend = 1,
             linetype = "dashed", colour = "firebrick", linewidth = 0.6) +
    scale_x_log10(
      breaks = c(1, 10, 100, 1000, 10000),
      labels = c("1", "10", "100", "1000", "10000")
    ) +
    scale_color_manual(
      values = mycols,
      labels = scales::label_parse(),
      name   = "Mode"
    ) +
    scale_linetype_manual(
      values = c("Empirical" = "solid", "Markov prediction" = "dashed"),
      name = NULL
    ) +
    labs(
      title = paste(specie),
      x = expression(tau),
      y = expression("|" * lambda[i]^tau * "|")
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    theme_classic(base_size = 13) +
    theme(
      plot.title       = element_text(face="bold", hjust=0.5),
      legend.position  = "right",
      panel.grid.major = element_line(color="grey92"),
      axis.line        = element_line(color="grey40")
    )
  
  print(p)
}
```

## 5.1. Length-Collapsed Version 

```{r 5.1}
#### functions ####

lagT <- function(seqs, tau) {
  
  states <- sort(unique(unlist(seqs)))
  n <- length(states)
  state_idx <- setNames(seq_along(states), states)
  
  mat <- matrix(0, nrow=n, ncol=n, dimnames=list(states, states))
  
  for (seq in seqs) {
    if (length(seq) <= tau) next
    
    #create transition matrix at lag tau
    for (i in 1:(length(seq) - tau)) {
      from <- as.character(seq[i])
      to   <- as.character(seq[i + tau])
      mat[from, to] <- mat[from, to] + 1
    }
  }
  
  #row-normalize (rows = "from" states)
  row_sums <- rowSums(mat)
  mat <- mat / ifelse(row_sums > 0, row_sums, 1)
  return(mat)
}

#first order markov tau is just T(1)^tau
matrixPower <- function(M, tau) {
  result <- diag(nrow(M))
  base <- M
  while (tau > 0) {
    if (tau %% 2 == 1) result <- result %*% base
    base <- base %*% base
    tau <- tau %/% 2
  }
  return(result)
}

#### visualizations ####

tau_values <- c(1, 2, 5, 10, 20, 50, 100, 200, 500, 1000)
n_modes <- 5  # plot top 5 non-trivial eigenvalues (mu = 2..6 in paper)

speciesList <- c('bre', 'mil', 'mul', 'oce', 'orn', 'pul')

colors <- paletteer_d("rcartocolor::Prism")
mycols <- colors[4:9]

for (specie in speciesList) {
  dat3 <- dat %>%
    filter(species == specie) %>%
    arrange(trial_id, sequence_id, fish_id, order_id)
  
  trueSeqs <- dat3 %>%
    group_by(trial_id, sequence_id, fish_id) %>%
    summarise(series = list(as.character(state)), .groups = "drop") %>%
    pull(series)
  
  #T(1) empirical
  P1 <- lagT(trueSeqs, tau = 1)
  
  #init
  eig_empirical <- matrix(NA, nrow=length(tau_values), ncol=n_modes)
  eig_markov    <- matrix(NA, nrow=length(tau_values), ncol=n_modes)
  
  for (k in seq_along(tau_values)) {
    tau <- tau_values[k]
    
    Ptau <- lagT(trueSeqs, tau)
    
    ev_emp <- sort(Mod(eigen(Ptau, only.values=TRUE)$values), decreasing=TRUE)
    eig_empirical[k, ] <- ev_emp[2:(n_modes+1)]
    
    ev_1 <- sort(Mod(eigen(P1, only.values=TRUE)$values), decreasing=TRUE)
    eig_markov[k, ] <- ev_1[2:(n_modes+1)]^tau
  }
  
  # Build tidy long-format data frames
  mode_labels <- paste0("mu==", 2:(n_modes+1))
  
  df_emp <- as.data.frame(eig_empirical) %>%
    setNames(mode_labels) %>%
    mutate(tau = tau_values) %>%
    pivot_longer(-tau, names_to="mode", values_to="lambda") %>%
    mutate(type = "Empirical")
  
  df_mkv <- as.data.frame(eig_markov) %>%
    setNames(mode_labels) %>%
    mutate(tau = tau_values) %>%
    pivot_longer(-tau, names_to="mode", values_to="lambda") %>%
    mutate(type = "Markov prediction")
  
  df_plot <- bind_rows(df_emp, df_mkv)
  
  p <- ggplot(df_plot, aes(x=tau, y=lambda, color=mode, linetype=type, group=interaction(mode, type))) +
    geom_line(linewidth=0.8) +
    geom_point(data = filter(df_plot, type == "Empirical"), size=2) +
    scale_x_log10(
      breaks = c(1, 10, 100, 1000),
      labels = c("1", "10", "100", "1000")
    ) +
    scale_color_manual(
      values = mycols,
      labels = scales::label_parse(),
      name   = "Mode"
    )+
    scale_linetype_manual(
      values = c("Empirical" = "solid", "Markov prediction" = "dashed"),
      name = NULL
    ) +
    labs(
      title = paste("Eigenvalue decay: ", specie),
      x = expression("log10(" * tau * ")"),
      y = expression("|" * lambda * "|")
    ) +
    coord_cartesian(ylim = c(0, 1)) +
    theme_classic(base_size = 13) +
    theme(
      plot.title    = element_text(face="bold", hjust=0.5),
      legend.position = "right",
      panel.grid.major = element_line(color="grey92"),
      axis.line     = element_line(color="grey40")
    )
  
  print(p)
}
```

# 6. Marginal State Distributions

```{r 6}
#### ── Transition Probability Stream Graph ────────────────────────────────── ####

library(tidyverse)
library(ggplot2)
library(patchwork)
library(paletteer)
library(scales)

#### ── Helpers ────────────────────────────────────────────────────────────── ####

lagT <- function(seqs, tau) {
  states   <- sort(unique(unlist(seqs)))
  n        <- length(states)
  mat      <- matrix(0, nrow = n, ncol = n, dimnames = list(states, states))
  for (seq in seqs) {
    if (length(seq) <= tau) next
    for (i in 1:(length(seq) - tau)) {
      from <- as.character(seq[i])
      to   <- as.character(seq[i + tau])
      mat[from, to] <- mat[from, to] + 1
    }
  }
  row_sums <- rowSums(mat)
  mat / ifelse(row_sums > 0, row_sums, 1)
}

make_ribbon_df <- function(df, state_order) {
  df %>%
    mutate(state = factor(state, levels = state_order)) %>%
    arrange(tau_log, state) %>%
    group_by(tau_log) %>%
    mutate(
      prob = prob / sum(prob),
      ymax = cumsum(prob),
      ymin = ymax - prob
    ) %>%
    ungroup()
}

#### ── Parameters ─────────────────────────────────────────────────────────── ####

tau_stream  <- c(1, 5, 10, 25, 50, 100, 250, 500, 1000, 2500, 5000, 7500, 10000, 15000)
speciesList <- c('bre', 'mil', 'mul', 'oce', 'orn', 'pul')

colors_all  <- paletteer_d("rcartocolor::Prism")
state_order <- c('a','b','c','d','e','f','g','h','j','k')
state_cols  <- setNames(colors_all[1:10], rev(state_order))

x_breaks <- log10(c(1, 10, 100, 1000, 10000))
x_labels <- c("1", "10", "100", "1,000", "10,000")

#### ── Main loop ──────────────────────────────────────────────────────────── ####

for (specie in speciesList) {

  cat("Processing:", specie, "\n")

  dat3 <- dat %>%
    filter(species == specie) %>%
    arrange(trial_id, sequence_id, fish_id, order_id)

  trueSeqs <- dat3 %>%
    group_by(trial_id, sequence_id, fish_id) %>%
    summarise(series = list(as.character(rep(state, length))), .groups = "drop") %>%
    pull(series)

  states <- sort(unique(unlist(trueSeqs)))

  # Q3 cutoff
  q3_len     <- quantile(sapply(trueSeqs, length), 0.75)
  cutoff_log <- log10(q3_len)

  ## ── 1. Baseline marginal (τ = 0) ───────────────────────────────────────── ##
  all_obs  <- unlist(trueSeqs)
  baseline <- as.numeric(table(factor(all_obs, levels = states)) / length(all_obs))
  names(baseline) <- states

  ## ── 2. Marginal P(state @ τ) ────────────────────────────────────────────── ##
  stream_rows <- map_dfr(tau_stream, function(tau) {
    Ptau <- lagT(trueSeqs, tau)
    tibble(
      tau     = tau,
      tau_log = log10(tau),
      state   = colnames(Ptau),
      prob    = as.numeric(baseline %*% Ptau)
    )
  })

  df_ribbon <- make_ribbon_df(stream_rows, state_order)

  ## ── 3. Baseline bar data ────────────────────────────────────────────────── ##
  baseline_df <- tibble(
    state = factor(states, levels = state_order),
    prob  = baseline[states]
  ) %>%
    arrange(state) %>%
    mutate(ymax = cumsum(prob), ymin = ymax - prob)

  ## ── 4. Baseline bar ─────────────────────────────────────────────────────── ##
  p_bar <- ggplot(baseline_df, aes(fill = state)) +
    geom_rect(aes(xmin = 0.1, xmax = 0.9, ymin = ymin, ymax = ymax),
              colour = "white", linewidth = 0.3) +
    geom_text(aes(x = 0.5, y = (ymin + ymax) / 2,
                  label = percent(prob, accuracy = 1)),
              size = 3, fontface = "bold", colour = "white") +
    scale_fill_manual(values = state_cols, guide = "none") +
    scale_x_continuous(limits = c(0, 1), breaks = NULL) +
    labs(x = NULL, y = "P(state)") +
    theme_classic(base_size = 11) +
    theme(
      axis.text.x  = element_blank(),
      axis.ticks.x = element_blank(),
      axis.line.x  = element_blank(),
      plot.margin  = margin(t = 5, r = 0, b = 5, l = 5),
      axis.ticks.y = element_blank(),
      axis.text.y  = element_blank()
    )

  ## ── 5. Ribbon plot ──────────────────────────────────────────────────────── ##
  p_stream <- ggplot(df_ribbon,
                     aes(x = tau_log, ymin = ymin, ymax = ymax, fill = state)) +
    geom_ribbon(colour = "white", linewidth = 0.25, alpha = 0.92) +
    annotate("rect",
             xmin = cutoff_log, xmax = log10(max(tau_stream)),
             ymin = 0, ymax = 1,
             fill = "white", alpha = 0.45) +
    annotate("segment",
             x = cutoff_log, xend = cutoff_log, y = 0, yend = 1,
             linetype = "dashed", colour = "firebrick", linewidth = 0.6) +
    scale_fill_manual(values = state_cols, name = "State",
                      breaks = rev(state_order)) +
    scale_x_continuous(breaks = x_breaks, labels = x_labels, expand = c(0.01, 0)) +
    labs(
      x     = expression(tau),
      y     = expression(P(state ~ "@" ~ tau)),
    ) +
    theme_classic(base_size = 11) +
    theme(
      plot.title      = element_text(face = "bold", hjust = 0.5, size = 12),
      legend.position = "right",
      axis.line       = element_line(colour = "grey40"),
      plot.margin     = margin(t = 5, r = 5, b = 5, l = 0),
      axis.ticks.y    = element_blank(),
      axis.text.y     = element_blank()
    )

  ## ── 6. Assemble ─────────────────────────────────────────────────────────── ##
  final_plot <- wrap_elements(full = p_bar) + wrap_elements(full = p_stream) +
    plot_layout(widths = c(1, 9)) +
    plot_annotation(
      title = specie,
      theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
    )

  print(final_plot)

}
```

```{r complicated}
#### ── Transition Probability Stream Graph + Companion Heatmap ────────────── ####

library(tidyverse)
library(ggplot2)
library(patchwork)
library(paletteer)
library(scales)

#### ── Helpers ────────────────────────────────────────────────────────────── ####

lagT <- function(seqs, tau) {
  states   <- sort(unique(unlist(seqs)))
  n        <- length(states)
  mat      <- matrix(0, nrow = n, ncol = n, dimnames = list(states, states))
  for (seq in seqs) {
    if (length(seq) <= tau) next
    for (i in 1:(length(seq) - tau)) {
      from <- as.character(seq[i])
      to   <- as.character(seq[i + tau])
      mat[from, to] <- mat[from, to] + 1
    }
  }
  row_sums <- rowSums(mat)
  mat / ifelse(row_sums > 0, row_sums, 1)
}

matrixPower <- function(M, tau) {
  result <- diag(nrow(M))
  base   <- M
  while (tau > 0) {
    if (tau %% 2 == 1) result <- result %*% base
    base <- base %*% base
    tau  <- tau %/% 2
  }
  result
}

make_ribbon_df <- function(df, state_order) {
  df %>%
    mutate(state = factor(state, levels = state_order)) %>%
    arrange(tau_log, state) %>%
    group_by(tau_log) %>%
    mutate(
      prob = prob / sum(prob),
      ymax = cumsum(prob),
      ymin = ymax - prob
    ) %>%
    ungroup()
}

lagT_coverage <- function(seqs, tau) {
  n_contributing <- sum(sapply(seqs, function(s) length(s) > tau))
  n_total        <- length(seqs)
  list(
    n_pairs       = sum(sapply(seqs, function(s) max(0, length(s) - tau))),
    n_seq         = n_contributing,
    coverage      = n_contributing / n_total
  )
}

#### ── Parameters ─────────────────────────────────────────────────────────── ####

tau_stream  <- c(1, 5, 10, 25, 50, 100, 250, 500, 1000, 2500, 5000, 7500, 10000, 15000)
tau_heat    <- c(1, 100, 1000, 5000, 10000)
speciesList <- c('bre', 'mil', 'mul', 'oce', 'orn', 'pul')

colors_all  <- paletteer_d("rcartocolor::Prism")
state_order <- c('a','b','c','d','e','f','g','h','j','k')
state_cols  <- setNames(colors_all[1:10], rev(state_order))

x_breaks <- log10(c(1, 10, 100, 1000, 10000))
x_labels <- c("1", "10", "100", "1,000", "10,000")

#### ── Main loop ──────────────────────────────────────────────────────────── ####

for (specie in speciesList) {

  cat("Processing:", specie, "\n")

  dat3 <- dat %>%
    filter(species == specie) %>%
    arrange(trial_id, sequence_id, fish_id, order_id)

  trueSeqs <- dat3 %>%
    group_by(trial_id, sequence_id, fish_id) %>%
    summarise(series = list(as.character(rep(state, length))), .groups = "drop") %>%
    pull(series)

  states <- sort(unique(unlist(trueSeqs)))

  ## ── 1. Baseline marginal (τ = 0) ───────────────────────────────────────── ##
  all_obs  <- unlist(trueSeqs)
  baseline <- as.numeric(table(factor(all_obs, levels = states)) / length(all_obs))
  names(baseline) <- states

  ## ── 2. Marginal P(state @ τ) ────────────────────────────────────────────── ##
  stream_rows <- map_dfr(tau_stream, function(tau) {
    Ptau <- lagT(trueSeqs, tau)
    tibble(
      tau     = tau,
      tau_log = log10(tau),
      state   = colnames(Ptau),
      prob    = as.numeric(baseline %*% Ptau)
    )
  })

  df_ribbon <- make_ribbon_df(stream_rows, state_order)
  
  # Build coverage data alongside stream_rows
  coverage_rows <- map_dfr(tau_stream, function(tau) {
    cv <- lagT_coverage(trueSeqs, tau)
    tibble(
      tau          = tau,
      tau_log      = log10(tau),
      n_pairs      = cv$n_pairs,
      n_seq        = cv$n_seq,
      coverage     = cv$coverage
    )
  })
  
  # Coverage threshold — grey out or truncate below this
  coverage_thresh <- 0.5

  ## ── 3. Baseline bar data ────────────────────────────────────────────────── ##
  baseline_df <- tibble(
    state = factor(states, levels = state_order),
    prob  = baseline[states]
  ) %>%
    arrange(state) %>%
    mutate(ymax = cumsum(prob), ymin = ymax - prob)

  ## ── 4. Heatmap data ─────────────────────────────────────────────────────── ##
  df_heat <- map_dfr(tau_heat, function(tau) {
    Ptau <- lagT(trueSeqs, tau)
    as.data.frame(as.table(Ptau)) %>%
      rename(from = Var1, to = Var2, prob = Freq) %>%
      mutate(tau = tau)
  }) %>%
    mutate(
      tau_label = paste0("tau == ", tau),
      tau_label = factor(tau_label, levels = paste0("tau == ", tau_heat))
    )

  ## ── 5. Baseline bar ─────────────────────────────────────────────────────── ##
p_bar <- ggplot(baseline_df, aes(fill = state)) +
    geom_rect(aes(xmin = 0.1, xmax = 0.9, ymin = ymin, ymax = ymax),
              colour = "white", linewidth = 0.3) +
    geom_text(aes(x = 0.5, y = (ymin + ymax) / 2,
                  label = percent(prob, accuracy = 1)),
              size = 3, fontface = "bold", colour = "white") +
    scale_fill_manual(values = state_cols, guide = "none") +
    scale_x_continuous(limits = c(0, 1), breaks = NULL) +
    #scale_y_continuous(limits = c(0, 1), expand = c(0, 0),
                       #labels = percent_format(accuracy = 1)) +
    labs(x = NULL, y = "P(state)") +
    theme_classic(base_size = 11) +
    theme(
      axis.text.x  = element_blank(),
      axis.ticks.x = element_blank(),
      axis.line.x  = element_blank(),
      plot.margin  = margin(t = 5, r = 0, b = 5, l = 5),
      axis.ticks.y       = element_blank(),
      axis.text.y = element_blank()
    )

  ## ── 6. Ribbon plot ──────────────────────────────────────────────────────── ##
  p_stream <- ggplot(df_ribbon,
                     aes(x = tau_log, ymin = ymin, ymax = ymax, fill = state)) +
    geom_ribbon(colour = "white", linewidth = 0.25, alpha = 0.92) +
    scale_fill_manual(values = state_cols, name = "State",
                      breaks = rev(state_order)) +
    scale_x_continuous(breaks = x_breaks, labels = x_labels, expand = c(0.01, 0)) +
    #scale_y_continuous(limits = c(0, 1), expand = c(0, 0),
                       #labels = percent_format(accuracy = 1)) +
    labs(
      x     = expression(tau),
      y     = expression(P(state ~ "@" ~ tau)),
      title = paste("Marginal State Probabilities over Time: ", specie)
    ) +
    theme_classic(base_size = 11) +
    theme(
      plot.title         = element_text(face = "bold", hjust = 0.5, size = 12),
      legend.position    = "right",
      #panel.grid.major.y = element_line(colour = "grey92"),
      axis.line          = element_line(colour = "grey40"),
      plot.margin        = margin(t = 5, r = 5, b = 5, l = 0),
      axis.ticks.y       = element_blank(),
      axis.text.y = element_blank()
    )
  
  #coverage plot
  p_coverage <- ggplot(coverage_rows, aes(x = tau_log, y = coverage)) +
  geom_area(fill = "grey70", alpha = 0.5) +
  geom_line(colour = "grey30", linewidth = 0.7) +
  geom_hline(yintercept = coverage_thresh, linetype = "dashed",
             colour = "firebrick", linewidth = 0.6) +
  annotate("text", x = min(coverage_rows$tau_log), y = coverage_thresh + 0.03,
           label = paste0(coverage_thresh * 100, "% threshold"),
           hjust = 0, size = 3, colour = "firebrick") +
  scale_x_continuous(breaks = x_breaks, labels = x_labels, expand = c(0.01, 0)) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0),
                     labels = percent_format(accuracy = 1)) +
  labs(x = expression(tau), y = "Seq. coverage") +
  theme_classic(base_size = 11) +
  theme(
    panel.grid.major.y = element_line(colour = "grey92"),
    axis.line          = element_line(colour = "grey40"),
    plot.margin        = margin(t = 0, r = 5, b = 5, l = 0)
  )
  
  cutoff_log <- coverage_rows %>%
  filter(coverage < coverage_thresh) %>%
  slice_min(tau_log, n = 1) %>%
  pull(tau_log)
  
  if (length(cutoff_log) > 0) {
  p_stream <- p_stream +
    annotate("rect",
             xmin = cutoff_log, xmax = max(coverage_rows$tau_log),
             ymin = 0, ymax = 1,
             fill = "white", alpha = 0.45) +
    annotate("segment",
             x = cutoff_log, xend = cutoff_log, y = 0, yend = 1,
             linetype = "dashed", colour = "firebrick", linewidth = 0.6)
  }
  
  p_combined <- wrap_elements(full = p_bar) + wrap_elements(full = p_stream) +
  plot_layout(widths = c(1, 9))



  ## ── 7. Heatmap ──────────────────────────────────────────────────────────── ##
  p_heat <- ggplot(df_heat,
                   aes(x = from, y = fct_rev(to), fill = prob)) +
    geom_tile(colour = "white", linewidth = 0.4) +
    # geom_text(
    #   aes(label = if_else(prob >= 0.05, sprintf("%.2f", prob), "")),
    #   size = 2.5, colour = "white", fontface = "bold"
    # ) +
    facet_wrap(~tau_label, nrow = 1, labeller = label_parsed) +
    scale_fill_paletteer_c("ggthemes::Red-Gold", name = "P(next @ τ|prev)") +
    labs(x = "Next state", y = "Prev state",
         title = paste("Conditional transition matrices: ", specie)) +
    theme_classic(base_size = 11) +
    theme(
      plot.title       = element_text(face = "bold", hjust = 0.5, size = 12),
      strip.background = element_rect(fill = "grey20", colour = NA),
      strip.text       = element_text(colour = "white", face = "bold", size = 10),
      axis.text        = element_text(size = 9),
      panel.spacing    = unit(0.4, "lines"),
      legend.position  = "right"
    )

  ## ── 8. Assemble with wrap_elements() ────────────────────────────────────── ##
  # wrap_elements() bypasses patchwork's add_guides() path that triggers
  # the Ops.data.frame bug in patchwork 1.1.3 + ggplot2 3.5.x.
  p_combined <- wrap_elements(full = p_bar) + wrap_elements(full = p_stream) +
    plot_layout(widths = c(1, 9))

  final_plot <- wrap_elements(full = p_combined) /
              wrap_elements(full = p_heat) /
              wrap_elements(full = p_coverage) +
  plot_layout(heights = c(2, 1.4, 0.6)) +
  plot_annotation(
    title = specie,
    theme = theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  )

  print(final_plot)

  # ggsave(paste0("transition_", specie, ".pdf"), final_plot, width = 14, height = 9)
}

#### ── Optional: anchor-conditioned small multiples ─────────────────────── ####
# anchor_ribbon_plot <- function(trueSeqs, anchor_state, tau_values,
#                                state_cols, state_order) {
#   states    <- sort(unique(unlist(trueSeqs)))
#   pi_anchor <- setNames(as.numeric(states == anchor_state), states)
#   rows <- map_dfr(tau_values, function(tau) {
#     Ptau <- lagT(trueSeqs, tau)
#     tibble(tau=tau, tau_log=log10(tau), state=colnames(Ptau),
#            prob=as.numeric(pi_anchor %*% Ptau))
#   })
#   df_r <- make_ribbon_df(rows, state_order)
#   ggplot(df_r, aes(x=tau_log, ymin=ymin, ymax=ymax, fill=state)) +
#     geom_ribbon(colour="white", linewidth=0.25, alpha=0.92) +
#     scale_fill_manual(values=state_cols, name="State", breaks=rev(state_order)) +
#     scale_x_continuous(breaks=log10(c(1,10,100,1000,10000)),
#                        labels=c("1","10","100","1,000","10,000")) +
#     scale_y_continuous(limits=c(0,1), expand=c(0,0),
#                        labels=percent_format(accuracy=1)) +
#     labs(title=paste0("P(state @ τ | start = '", anchor_state, "')"),
#          x=expression(tau), y="Probability") +
#     theme_classic(base_size=11)
# }
# sm_plots <- map(states, ~anchor_ribbon_plot(trueSeqs, .x, tau_stream,
#                                              state_cols, state_order))
# wrap_plots(sm_plots, ncol=5) + plot_layout(guides="collect")
```



